Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/intro.tex
Writing from last night

\end{document}

\add{To return to the context of this thesis, the observed data $y$ is an
estimated transmission tree for a viral epidemic under investigation. The model
in question is a contact network model with parameters $\theta$. $z$ is a
simulated transmission tree, obtained by first generating a contact network
under the model, and then simulating the spread of an epidemic over that
network. A transmission tree can be constructed by keeping track of who
infected whom during the simulated epidemic (further details will be given in
\cref{subsec:nettree}). The distance function $\rho$ must compare two
transmission trees - the observed tree $y$, and the simulated tree $z$. We will
define this distance function using the tree kernel discussed in
\cref{subsec:treeshape}. In words, the approximate posterior we consider here
is a distribution which assigns a joint probability density to model parameters
and simulated transmission trees under those parameters. The probability
density is proportional to the product of the prior on the parameters, and the
likelihood of the simulated transmission tree under those parameters, but only
if the simulated transmission tree is sufficiently close to the true tree.
Otherwise, the probability density is zero.}

{\color{red}\sout{As we shall see in the next section, this distribution can
be sampled from exactly. The word ``approximate'' derives from assumption that,
for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
marginal in $z$ of this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is, }}
{\color{blue}\uline{In fact, it is not the \gls{ABC} target distribution
itself, but rather its marginal in $z$, which approximates the posterior
distribution. In other words, we claim that }} 
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \Pr(\theta \mid y).
\]
The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

@@

In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  \Pr(\theta \mid y) = \Pr(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
A number of sophisticated methods have been developed for selecting and
weighting summary statistics based on various optimality criteria}
\autocite[][and references therein]{aeschbacher2012novel,
blum2013comparative}}.

Summary statistics can be useful if the data are high-dimensional or of a
complex type, but they are not strictly necessary. For instance, if the data
are numeric and of low dimension, the distance function may simply be the
Euclidean distance~\autocite{sisson2007sequential}.
{\color{blue}\textcite{park2015k2} \uline{proposed the use of a kernel function}
(defined in \cref{subsec:treeshape}) \uline{in place of a distance function.
The authors referred to their approach as ``double-kernel \gls{ABC}'' due to
the use of a second kernel function to compute the weights of the particles.
The work by \textcite{poon2015phylodynamic}, upon which ours is based, employed
a similar approach, replacing the likelihood ratio in Bayesian \gls{MCMC} with
a ratio of kernel scores.}}

@@

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
simplify the notation, we shall restrict the descriptions of these algorithms
to the case of one simulated dataset per parameter particle (the meaning of
this will become clear shortly). The extension to multiple datasets per
particle is straightforward and will be given at the end of the section. We use
the variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC}
target distribution can be written $\pi_\varepsilon(x \mid y)$.

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly, tavare1997inferring}. The algorithm,
outlined in \cref{alg:abcrej}, repeats the following steps until a desired
number of samples from the target distribution are obtained. Parameter values
$\theta$ are sampled according to the prior distribution $\pi(\theta)$. Then, a
simulated dataset $z$ is generated from the model with the sampled parameter
values. By definition, the probability density of obtaining the particular
dataset $z$ is $f(z \mid \theta)$. Finally, the parameters are sampled if the
distance of $z$ from the observed data $y$ is less than $\varepsilon$, that is,
with probability $\I_{A_{\varepsilon, y}}(z)$. Putting this all together, the
parameters $\theta$ are sampled with probability proportional to
\[
  \pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}}(z),
\]
which is exactly the numerator of the \gls{ABC} target distribution. Thus,
$\theta$ represents an unbiased sample from the approximate posterior.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i(x) = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of
tolerances $\varepsilon_i$. The intention is for the algorithm to progress
smoothly through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. {\color{blue}\uline{The initial value
$\varepsilon_0$ is set to $\infty$, which makes the first distribution in the
sequence}
\[
  \pi_0(\theta, z) = \frac{\Pr(\theta) \Pr(z \mid \theta)}
               {\int_{\R \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d\theta \d z}.
\]
\uline{For brevity, we have ommitted the dependence on the observed data $y$.
In the terminology of} \cref{alg:smc}, \uline{the numerator is the first of the
$\gamma$'s, that is, $\gamma_0 = \Pr(\theta) \Pr(z \mid \theta)$. Sampling in
proportion to $\gamma_0$ is straightforward and was already demonstrated for
rejection \gls{ABC} above. Because the sampling is exact, the initial
importance weights are all set equal to 1 and normalized to $1/n$ where $n$ is
the number of particles.}}

As discussed in \cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$
is problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
Several options have been proposed~\autocite{beaumont2009adaptive,
sisson2007sequential, del2012adaptive}. {\color{blue}\uline{With an
appropriately chosen kernel, the weight update} (\cref{eq:smcwt}) \uline{will
simplify into a computable expression.} \textcite{sisson2007sequential} and
\textcite{beaumont2009adaptive} \uline{both suggest random walk kernels for
$K$, where each particle is
perturbed according to a Gaussian distribution. The backwards kernels $L$ are
chosen to approximately minimize the variance in importance weights. In this
thesis, we use an \gls{MCMC} kernel, as proposed by} \textcite{del2012adaptive}.
\uline{The associated backwards kernels and weight updates are discussed in the
next chapter} (\cref{subsec:adaptsmc}). \uline{With generic kernels $K$ and
$L$, the \gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC}
sampler} (\cref{alg:smcsamp}), \uline{with $\gamma_i$ replaced with $\pi_i$.}
Therefore we will not repeat the full algorithm here.}

@@

{\color{blue}\uline{In} \cref{sec:obj}, \uline{we outlined three research
objectives that will be addressed in this thesis. First, we aim to develop a
method for fitting contact network models to estimated transmission trees.
Although transmission trees can be estimated for any epidemic by thorough
contact tracing, viral diseases are the most useful context for this method.
Transmission and sequence evolution occur on similar time scales for RNA
viruses, resulting in viral phylogenies whose shapes are heavily constrained by
the transmission process. The study of this interaction between evolution and
epidemiology is called \emph{phylodynamics}. Phylodynamic methods make it
possible to estimate transmission trees from viral sequence data; these
estimated trees form the input data for our method.}

\uline{From these estimated transmission trees, our method's purpose is to
estimate the parameters of a contact network model. Rather than assuming a
homogeneously mixed population, as most epidemiological models do, network
models take the more realistic view that human populations are structured. That
is, contacts which allow for transmission may occur in a nonrandom way, rather
than between every pair of individuals in the population. A network model
parameterizes this structure. In particular, our second research objective is
to characterize our ability to fit the \gls{BA} model, which incorporates
preferential attachment to generate networks with realistic degree
distributions.}

\uline{Our method will apply \gls{ABC}, a simulation-based approach, for
fitting these models. Simulating a transmission tree according to a network
model is straightforward: a network can be generated according to the model,
and the spread of an epidemic can be simulated over the network and recorded in
a transmission tree. \gls{ABC} uses the concordance between these simulated
transmission trees and the ``true'' estimated tree as an indicator of parameter
credibility. The closer the simulated transmission trees appear to the true
tree, the more weight is assigned to the associated network parameters.}

\uline{\gls{ABC} can be implemented by at least three algorithms, but the one
we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a population of
parameter ``particles'' to approximate a distribution of interest, in this case
the approximate posterior distribution targeted by \gls{ABC}. After running the
\gls{ABC}-\gls{SMC} algorithm, statistics on the model parameters can be
approximated by the weighted population of particles. For example, a weighted
average would give an approximate expected value for each parameter.}

\uline{Thus, our method integrates four research topics: phylogenetics, contact
networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
the problem domain. Phylogenetic data is the input to our method, while
estimates of the parameters of contact network models are the desired output.
The latter two topics define the algorithm and statistical framework that our
inference method will use. }}

>>>>

\add{To return to the context of this thesis, the observed data $y$ is an
estimated transmission tree for a viral epidemic under investigation. The model
in question is a contact network model with parameters $\theta$. The simulated
dataset $z$ is a transmission tree, obtained by first generating a contact
network under the model, and then simulating the spread of an epidemic over
that network. A transmission tree can be constructed by keeping track of who
infected whom during the simulated epidemic (further details will be given in
\cref{subsec:nettree}). The distance function $\rho$ must compare two
transmission trees - the observed tree, which takes the place of $y$, and the
simulated tree $z$. We will define this distance function using the tree kernel
discussed in \cref{subsec:treeshape}. In words, the approximate posterior we
consider here is a distribution which assigns a joint probability density to
model parameters and simulated transmission trees under those parameters. The
probability density is proportional to the product of the prior on the
parameters, and the likelihood of the simulated transmission tree under those
parameters, but only if the simulated transmission tree is sufficiently close
to the true tree.  Otherwise, the probability density is zero.}

\del{As we shall see in the next section, this distribution can be sampled from
exactly. The word ``approximate'' derives from assumption that, for a suitably
chosen distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,}
\add{In fact, it is not the \gls{ABC} target distribution
itself, but rather its marginal in $z$, which approximates the posterior
distribution. In other words, we claim that}
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \post(\theta \mid y).
\]
The intuition for why this approximation might be reasonable comes from
\add{considering the case when $\varepsilon$ = 0 and $\rho$ has the property
that $\rho(z, y)$ if and only if $x = y$. In that case,} \del{the fact that,
when $\varepsilon = 0$,} the $\varepsilon$-ball around $y$ should contain only
$y$ itself, hence the integral on the left is exactly equal to the posterior.
Thus, by taking $\varepsilon$ small, we should attain something close to the
posterior \add{if $\rho$ captures the similarity between datasets reasonably
well. However, the accuracy of the \gls{ABC} approximation depends heavily on
the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}

@@

In many applications \add{(eg. \autocite{fu1997estimating,tanaka2006using})},
$\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which
maps data points into a vector of summary statistics.
\add{In the context of \gls{ABC}, a summary statistic $S$ is called
\defn{sufficient} if}
\[
  \post(\theta \mid y) = \post(\theta \mid S(y)).
\]
\add{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
A number of sophisticated methods have been developed for selecting and
weighting summary statistics based on various optimality criteria
\autocite[][and~references~therein]{aeschbacher2012novel,blum2013comparative}.
We do not apply these methods in this work, instead focusing on a distance
function which is not based on summary statistics.}

\del{Summary statistics can be useful if the data are high-dimensional or of a
complex type, but}\add{Although summary statistics are often presented as a
fundamental part of \gls{ABC},} they are not strictly necessary. For instance,
if the data are numeric and of low dimension, the distance function may simply
be the Euclidean distance~\autocite{sisson2007sequential}.
\add{\textcite{park2015k2} proposed the use of a kernel function
(as defined in \cref{subsec:treeshape}) in place of a distance function.  The
authors referred to their approach as ``double-kernel \gls{ABC}'' due to the
use of a second (unrelated) kernel function to compute the weights of the
particles. The work by \textcite{poon2015phylodynamic}, upon which ours is
based, employed a similar approach, replacing the likelihood ratio in Bayesian
\gls{MCMC} with a ratio of kernel scores.}

@@

Algorithms for performing \gls{ABC} can be grouped into three categories:
rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
simplify the notation, we shall restrict the descriptions of these algorithms
to the case of one simulated dataset per parameter particle (the meaning of
this will become clear shortly). The extension to multiple datasets per
particle is straightforward and will be given at the end of the section. We use
the variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC}
target distribution can be written $\pi_\varepsilon(x \mid y)$. \add{This makes
our notation consistent with \cref{sec:smc}.}

Rejection ABC is the simplest method, and also the one which was first
proposed~\autocite{rubin1984bayesianly, tavare1997inferring}. The algorithm,
outlined in \cref{alg:abcrej}, repeats the following steps until a desired
number of samples from the target distribution are obtained. Parameter values
$\theta$ are sampled according to the prior distribution $\pi(\theta)$. Then, a
simulated dataset $z$ is generated from the model with the sampled parameter
values. By definition, the probability density of obtaining the particular
dataset $z$ is $f(z \mid \theta)$. Finally, the parameters are sampled if the
distance of $z$ from the observed data $y$ is less than $\varepsilon$, that is,
with probability $\I_{A_{\varepsilon, y}}(z)$. Putting this all together, the
parameters $\theta$ are sampled with probability proportional to
\[
  \prior(\theta) \lik(z \mid \theta) \I_{A_{\varepsilon, y}}(z),
\]
which is exactly the numerator of the \gls{ABC} target distribution. Thus,
$\theta$ represents an unbiased sample from the approximate posterior.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i(x) = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of
tolerances $\varepsilon_i$. The intention is for the algorithm to progress
smoothly through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. \add{The initial value $\varepsilon_1$ is set
to $\infty$, which makes the first distribution in the sequence}
\[
  \madd{\pi_1(\theta, z) = \frac{\prior(\theta) \lik(z \mid \theta)}
  {\int_{\R \times \Theta} \prior(\theta) \lik(z \mid \theta) \d\theta \d z}}.
\]
\add{This initial distribution does not depend on the observed data $y$.  In
the terminology of the \gls{SMC} sampler (\cref{alg:smc}), the numerator is the
first of the $\gamma$'s, that is, $\gamma_1 = \prior(\theta) \lik(z \mid
\theta)$. Sampling in proportion to $\gamma_1$ is straightforward and was
already demonstrated for rejection \gls{ABC} above. Because the sampling is
exact, the initial importance weights are all set equal to 1 and normalized to
$1/\gls{n}$ where \gls{n} is the number of particles.}

As discussed in \cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$
is problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
Several options have been proposed~\autocite{beaumont2009adaptive,
sisson2007sequential, del2012adaptive}. \add{With an appropriately chosen
kernel, the weight update (\cref{eq:smcwt}) will simplify into a computable
expression. \textcite{sisson2007sequential} and \textcite{beaumont2009adaptive}
both suggest random walk kernels for $K$, where each particle is perturbed
according to a Gaussian distribution. The backwards kernels $L$ are chosen to
approximately minimize the variance in importance weights. In this thesis, we
use an \gls{MCMC} kernel, as proposed by \textcite{del2012adaptive}.  The
associated backwards kernels and weight updates are discussed in the next
chapter (\cref{subsec:adaptsmc}). With generic kernels $K$ and $L$, the
\gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC} sampler
(\cref{alg:smcsamp}), with $\gamma_i$ replaced with $\pi_i$.  Therefore we will
not repeat the full algorithm here.}

@@

\add{In \cref{sec:obj}, we outlined three research objectives that will be
addressed in this thesis. First, we aim to develop a method for fitting contact
network models to estimated transmission trees. Although transmission trees
can be estimated for any epidemic by thorough contact tracing, viral diseases
are the most useful context for this method due to the possibility of inferring
the trees from sequence data. Transmission and sequence evolution occur on
similar time scales for RNA viruses, resulting in viral phylogenies whose
shapes are heavily constrained by the transmission process. The study of this
interaction between evolution and epidemiology is called \emph{phylodynamics};
phylodynamic methods make it possible to estimate transmission trees from viral
sequence data. These estimated trees form the input data for our method.}

\add{The desired output of our method is a posterior distribution of the
parameters of a contact network model. Rather than assuming a homogeneously
mixed population, as most epidemiological models do, network models take the
more realistic view that human populations are structured. That is, contacts
which allow for transmission may occur in a nonrandom way, rather than between
every pair of individuals in the population. A network model parameterizes this
structure. In particular, our second research objective is to characterize our
ability to fit the \gls{BA} model, which incorporates preferential attachment
to generate networks with realistic degree distributions.}

\add{To fit these models, our method will apply \gls{ABC}, a simulation-based
approach. Simulating a transmission tree according to a network model is
straightforward: a network can be generated according to the model, and the
spread of an epidemic can be simulated over the network and recorded in a
transmission tree. \Gls{ABC} uses the concordance between these simulated
transmission trees and the ``true'' estimated tree as an indicator of parameter
credibility. The closer the simulated transmission trees appear to the true
tree, the more weight is assigned to the associated network parameters.}

\add{\Gls{ABC} can be implemented by at least three classes of algorithm, but
the one we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a
population of parameter ``particles'' to approximate a distribution of
interest, in this case the approximate posterior distribution targeted by
\gls{ABC}. After running the \gls{ABC}-\gls{SMC} algorithm, statistics on the
model parameters can be approximated by the weighted population of particles.
For example, a weighted average would give an approximate expected value for
each parameter.}

\add{Thus, our method integrates four research topics: phylogenetics, contact
networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
the problem domain. Phylogenetic data is the input to our method, while
estimates of the parameters of contact network models are the desired output.
The latter two topics define the algorithm and statistical framework that our
inference method will use.}
